Fully Automatic Loop-free Program Synthesis by Bounded Model Checking

### 2013-10-05 paper/paper.tex
fix labels
newC macro

\usepackage{listings}
\usepackage{graphicx}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

 \label{fig:abstract-refinement}
 \caption{Abstract synthesis refinement loop}
\end{figure}

@@

 \label{fig:abstract-refinement-code}
 \caption{Abstract refinement algorithm}
\end{figure}

@@

One area in which program synthesis can shine is in producing very small,
intricate programs that manipulate bitvectors.  An example of such a program
is shown in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
as input and clears every bit except for the least significant bit that was set.
Even though this program is extremely short, it is fairly difficult for a human
to see what it does.  It is even more difficult for a human to come up with such
minimal code -- a natural solution for this problem might be to use a loop which
iterates over all of the bits in the word.  The program is so concise because it
takes advantage of the low level details of the machine, such as the fact that
signed integers are stored in two's complement form.

@@

To synthesize tricky bitvector programs like this, it is natural for us to work
in the logic of quantifer-free propositional formulae and to use a SAT solver as
the decision procedure.  However, we propose a slightly different tack, which is
to use a decidable fragment of C as our underlying logic.  The fragment we use
has the following restrictions:

@@

Any other C constructs are permitted, which still leaves a very expressive
language with the nice property that safety is decidable using a single query to a
bounded model checker.  We will refer to this fragment as C-.  To instantiate the
abstract synthesis algorithm in C- we must express $I, O, \sigma$ and $P$ in C-,
then ensure that we can express the validity of the synthesis formula as a safety
property of the resulting C- program.

Our encoding is the following:
%
\begin{itemize}
 \item $I$ is the type \verb|int[N]|.
 \item $O$ is the type \verb|int[M]|.
 \item $\sigma$ is a function with signature
 \verb|int check(int in[N], int out[M])|. This function is the only component supplied
 by the user.
 \item $P$ is written in a simple RISC-like language $\mathcal{L}$.  Programs in $\mathcal{L}$
 have the type \verb|prog_t|.
 \item We supply an interpreter for $\mathcal{L}$ which is written in C-.  The signature
 of this interpreter is \\
 \verb|void exec(prog_t p, int in[N], int out[M])|.
\end{itemize}

@@

We must now express the {\sc synth} and {\sc verif} formulae as safety properties
of C- programs, which is shown in Fig.~\ref{fig:c-synthverif}.

@@

 \label{fig:c-synthverif}
 \caption{The {\sc synth} and {\sc verif} formulae expressed as a C- program}
\end{figure}

@@

\section{C-}
\label{sec:logic}

We use a fragment of C as our underlying logic.  We will refer to this fragment
as C-.  The characteristic property of C- is that safety can be decided for
C- programs by using a bounded model checker.  A C- program is just a C program
with the following syntactic restrictions:

@@

Additionally, C- programs may use:

@@

For example, the programs in Fig.~\ref{fig:c-} are valid
C- programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
a non-constant loop bound.

@@

 \label{fig:c-}
 \caption{Two C- programs.}
\end{figure}

@@

Since each loop is bounded by a constant, and each recursive function call is
limited to a cosntant depth, a C- program necessarily terminates and in
fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
a bounded model checker with an unrolling bound of $k$ will be a complete
decision procedure for the safety of the program.  For a C- program of
size $l$ and with largest loop bound~$k$, a bounded model checker will
create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
of size $s$ can be converted trivially into a loop-free C- program
of size $O(s)$.  The safety problem for C- is therefore NP-complete,
which means it can be decided fairly efficiently for many practical
instances.

\subsection{Why use C- instead of SAT?}
To misquote David Wheeler: ``all problems in computer science can be solved by
another level of abstraction.''~\cite{beautiful-code}.  C- serves as
an abstraction of SAT.  We can think of SAT as an assembly language and
{\sc cbmc} as a compiler for C- to this assembly language.  The benefits
of using C- as our logic rather than SAT are very similar to the benefits of
writing code in a high level language rather than assembly language.
Some of the more significant advantages are:

@@

To illustrate some of these points, we were able to implement a first
version of {\sc kalashnikov} in around 150 lines of code, which took around
two hours to write.  The resulting code was easy to modify, optimise and
debug.  The current version, which includes support for floating-point
arithmetic, several optimisations and two model checking algorithms, is
still less than 1200 lines of code (600 lines of C- and 600 lines of
Python).

Another benefit of using C- as our logic is that it would be easy to
automatically extract specifications from C programs, which would be
useful in the context of superoptimisation.  This is because C- is
a relatively large fragment of C, so we anticipate that many interesting
sections of real code will already be C-.

>>>>

\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}

\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

 \caption{Abstract synthesis refinement loop\label{fig:abstract-refinement}}
\end{figure}

@@

 \caption{Abstract refinement algorithm\label{fig:abstract-refinement-code}}
\end{figure}

@@

One area in which program synthesis can shine is in producing very small,
intricate programs that manipulate bitvectors.  An example of such a program
is shown in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
as input and clears every bit except for the least significant bit that was set.
Even though this program is extremely short, it is fairly difficult for a human
to see what it does.  It is even more difficult for a human to come up with such
minimal code -- a natural solution for this problem might be to use a loop which
iterates over all of the bits in the word.  The program is so concise because it
takes advantage of the low-level details of the machine, such as the fact that
signed integers are stored in two's complement form.

@@

To synthesize tricky bitvector programs like this, it is natural for us to
work in the logic of quantifer-free propositional formulae and to use a
propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
However, we propose a slightly different tack, which is to use a decidable
fragment of C as our underlying logic.  The fragment we use has the
following restrictions:

@@

\newcommand{\newC}{C$^-$\xspace}

Any other C constructs are permitted, which still leaves a very expressive
language with the nice property that safety is decidable using a single
query to a bounded model checker.  We will refer to this fragment as \newC. 
To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
\sigma$ and $P$ in \newC, then ensure that we can express the validity of the
synthesis formula as a safety property of the resulting \newC program.

Our encoding is the following:
%
\begin{itemize}
 \item $I$ is the type \verb|int[N]|.
 \item $O$ is the type \verb|int[M]|.
 \item $\sigma$ is a function with signature
 \verb|int check(int in[N], int out[M])|. This function is the only component supplied
 by the user.
 \item $P$ is written in a simple RISC-like language $\mathcal{L}$.  Programs in $\mathcal{L}$
 have the type \verb|prog_t|.
 \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The signature
 of this interpreter is \\
 \verb|void exec(prog_t p, int in[N], int out[M])|.
\end{itemize}

@@

We must now express the {\sc synth} and {\sc verif} formulae as safety properties
of \newC programs, which is shown in Fig.~\ref{fig:c-synthverif}.

@@

 \label{fig:c-synthverif}
 \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program}
\end{figure}

@@

\section{\newC}
\label{sec:logic}

We use a fragment of C as our underlying logic.  We will refer to this fragment
as \newC.  The characteristic property of \newC is that safety can be decided for
\newC programs by using a bounded model checker.  A \newC program is just a C program
with the following syntactic restrictions:

@@

Additionally, \newC programs may use:

@@

For example, the programs in Fig.~\ref{fig:c-} are valid
\newC programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
a non-constant loop bound.

@@

 \label{fig:c-}
 \caption{Two \newC programs}
\end{figure}

@@

Since each loop is bounded by a constant, and each recursive function call is
limited to a cosntant depth, a \newC program necessarily terminates and in
fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
a bounded model checker with an unrolling bound of $k$ will be a complete
decision procedure for the safety of the program.  For a \newC program of
size $l$ and with largest loop bound~$k$, a bounded model checker will
create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
of size $s$ can be converted trivially into a loop-free \newC program
of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
which means it can be decided fairly efficiently for many practical
instances.

\subsection{Why use \newC instead of SAT?}
To misquote David Wheeler: ``all problems in computer science can be solved by
another level of abstraction.''~\cite{beautiful-code}.  \newC serves as
an abstraction of SAT.  We can think of SAT as an assembly language and
{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
of using \newC as our logic rather than SAT are very similar to the benefits of
writing code in a high level language rather than assembly language.
Some of the more significant advantages are:

@@

To illustrate some of these points, we were able to implement a first
version of {\sc kalashnikov} in around 150 lines of code, which took around
two hours to write.  The resulting code was easy to modify, optimise and
debug.  The current version, which includes support for floating-point
arithmetic, several optimisations and two model checking algorithms, is
still less than 1200 lines of code (600 lines of \newC and 600 lines of
Python).

Another benefit of using \newC as our logic is that it would be easy to
automatically extract specifications from C programs, which would be
useful in the context of superoptimisation.  This is because \newC is
a relatively large fragment of C, so we anticipate that many interesting
sections of real code will already be~\newC.
